---
title: "Analysis_medal_per_athlete"
author: "stewarta"
date: "14 January 2018"
output: html_document
---

# R Analysing medals won per athlete/per country

Today I am introducing a sneaky little data analysis using [R](https://www.r-project.org/) on figure skating in the olympics. I am first going to look at the medals won using data from the [official Olympics database website](http://www.olympiandatabase.com)

## Looking at the total number of athletes that have won at least a medal per country

```{r fig.height = 5, fig.width= 15}
#loading libraries
library(ggplot2)

#reading csv file
medalCount <- read.csv(file="./stats_athletes_with_medals_count.csv", header=TRUE, sep=",")
medalCountdf <- data.frame(medalCount)
#number of athletes per country that have won at least one medal
tbl_total_num_athletes_with_medal <- table(medalCountdf$NationID)
barplot(tbl_total_num_athletes_with_medal, main="Number of medals won per country")
```
No surprises there, America has always entered as many athletes as they were possibly allowed and have the budget to bring them all.

But let's see if they still had the highest number of wins:

## Number of gold medals per country

```{r fig.height = 5, fig.width= 15}
tbl_gold <- table(medalCountdf[which(medalCountdf$G!=0),]$NationID)
barplot(tbl_gold, main="Number of gold medals won per country")

```

## Athletes that have won multiple times

```{r fig.height = 5, fig.width= 20}
multiple_gold_athletes <- medalCountdf[which(medalCountdf$G>1),]
print(multiple_gold_athletes)
#ggplot(data=multiple_gold_athletes, aes(x=Athlete, y=multiple_gold_athletes$G)) +
  #geom_bar(stat="identity")

#barplot(tbl_gold_athletes, main="Athletes with multiple wins")

```

## French athletes with medals

Since I am french and I used to skate in France, I'm very interested in how we did over the years so here are all the french athletes who have won at least a medal in the order they are scored according to how many and which color their medals were.
```{r fig.height = 5, fig.width= 20}
french_athletes_score <- medalCountdf[which(medalCountdf$NationID=="FRA"),]
print(french_athletes_score)

```

```{r fig.height = 5, fig.width= 20}
nations_with_athletes_with_no_wins <- table(medalCountdf[which(medalCountdf$G<1),]$NationID)
barplot(nations_with_athletes_with_no_wins, main="Number of athletes per country who have never won")

nations_with_athletes_with_no_wins <- medalCountdf[which(medalCountdf$G<1),]

```
